
---
title: "Har_Model_Botta"
author: 
  name: Manuel Schreiber
output: html_document

---



## Section 1: data preparation ##

$$ log(P_2) - log(P_2) = log(\frac{P2}{P1})$$ : "continuous grow rate from P1 to P2"
f.ex. log(96/95)=log(1.01053)~0.0105, i.e. a growth rate of 1.05% or the log return
the percentage change would be $$\frac{P2-P1}{P1}=\frac{96-95}{95}~0.0105$$

```{r}
# install libraries
if (!require("quantmod")) install.packages("quantmod")

# Load required libraries
suppressMessages({ library(quantmod); library(xts)
})
# Download S&P 500 data (2020-2023, i.e. 4 years)
getSymbols("^GSPC", from="2020-01-01", to="2023-12-31", auto.assign=TRUE, warnings=FALSE)

prices <- Cl(GSPC) # closing prices (252*4 = 1008)

print(length(prices))

# log(P_t) - log(P_{t-1}) = approximation of a percentage change (for small changes)
# continuously compounded rate of return (i.e. the log return)
returns <- diff(log(prices))[-1]  

print(returns)
print(prices)
```

```{r}
# Calculate realized variance (proxy using squared returns)
rv <- as.numeric(returns^2 * 252) # Annualized realized variance 

rv <- rv[!is.na(rv)] # Remove NAs

head(rv)

print( length(rv) )

# Create HAR regressors
n <- length(rv) 

start_idx <- 23

# Pre-allocate matrices
n_obs <- n - start_idx + 1

y <- numeric(n_obs); X <- matrix(NA, n_obs, 4)

length(y)

colnames(X) <- c("Intercept", "Daily", "Weekly", "Monthly")
```

```{r}
# Fill data matrices
for(i in 1:n_obs) {

t <- start_idx + i - 1
# Dependent variable
y[i] <- rv[t]
# Regressors
X[i, 1] <- 1 # intercept
X[i, 2] <- rv[t-1] # daily rv
X[i, 3] <- mean(rv[(t-5):(t-1)]) # weekly realized variance (5 trading days)
X[i, 4] <- mean(rv[(t-22):(t-1)])  # monthly rv

}

T_obs <- length(y)

dim(X) # shape of the regressor matrix

length(y) # length of the dep var vector

head(y) # dependent variable

head(X) # regressor matrix
```

## Section 2: MCMC functions ##

```{r}
# Gibbs sampler for Bayesian HAR
gibbs_har <- function(y, 
X, 
n_iter = 10000, 
burnin = 2000
) 
{ # Prior parameters
beta_0 <- c(0, 0.4, 0.3, 0.2) # Prior means
V_0 <- diag(c(100, 0.04, 0.04, 0.04)) # Prior covariance 
a_0 <- 2; b_0 <- 1 # InvGamma parameters for sigma^2

a_0

# Initialize
k <- ncol(X); T_obs <- length(y); beta <- beta_0; sigma2 <- 1
# Store only post-burnin samples
n_keep <- n_iter - burnin
beta_draws <- matrix(NA, n_keep, k); sigma2_draws <- numeric(n_keep)

# Precompute
V_0_inv <- solve(V_0); XtX <- t(X) %*% X; Xty <- t(X) %*% y

##print(V_0_inv)

for (iter in 1:n_iter) {
# Update beta | sigma2, y
V_n_inv <- V_0_inv + XtX / sigma2
V_n <- solve(V_n_inv)
beta_n <- V_n %*% (V_0_inv %*% beta_0 + Xty / sigma2) 
beta <- drop(beta_n + t(chol(V_n)) %*% rnorm(k))
# Update sigma2 | beta, y
residuals <- y - X %*% beta
a_n <- a_0 + T_obs / 2
b_n <- b_0 + sum(residuals^2) / 2 
sigma2 <- 1 / rgamma(1, a_n, b_n)

# Store post-burnin draws only
if (iter > burnin) {
beta_draws[iter - burnin, ] <- beta; sigma2_draws[iter - burnin] <- sigma2
  } # end if function

 } # end for loop

return(list(beta = beta_draws,
              sigma2 = sigma2_draws))

} # end function body
```

## Section 3: extracting the results ##

```{r}
results <- gibbs_har(y, X)

colMeans(results$beta) # Means of the intercept col, daily, weekly, monthly rv

mean(results$sigma2) # posterior mean for the std dev


##hist(results$beta)
```

## Section 4: MCMC diagnostics

Recap of the Markov Chain Monte Carlo algorithm:
* a sequence of random variables $\{X_t\}$ is produced
* this sequence is a Markov Chain
* the larger $t$ is, the more similar the distribution of $X_t$ is to the target distribution
* two terms in the chain, $X_t$ and $X_{t+n}, become almost independent as $n$ gets large 
* when we run a MCMC algorithm for $T$ periods, we obtain a MCMC sample of the first $T$ realizations of the chain, $$X_1,\ldots,X_T$$
* we then use the empirical distribution of this MCMC sample to approximate the target distribution

Two main problems arise:
* (P1) A large portion of the sample is drawn from distributions that are significantly different from the target distribution
* (P2) The effective size of the sample is too small

(P1) can be addressed by discarding a large chunk of initial observations (burn-in sample) or by increasing the no. of iterations (MCMC sample size)

Markov Chain Monte Carlo (MCMC) diagnostics are tools that can be used to check whether the quality of a sample generated with an MCMC algorithm is sufficient to provide an accurate approximation of the target distribution.

In particular, MCMC diagnostics are used to check:

* whether a large portion of the MCMC sample has been drawn from distributions that are significantly different from the target distribution
* whether the size of the generated sample is too small.

Diagnostic tools:
* The simplest way to diagnose problems in an MCMC sample is to split the sample into two or more chunks and check whether we get the same results on all the chunks.
* Another simple way to diagnose problems is to run the MCMC algorithm more than once with different (possibly, very different) starting points $x_{1}$, so as to obtain multiple MCMC samples.
* Trace plots may help us to understand what kind of problem (if any) affects our MCMC sample

# 4.1 Trace plots

```{r}
# Example for the second Beta parameter (weekly)
plot(results$beta[, 2], type = "l", col = "steelblue", 
     main = "Trace Plot for Beta_1 (daily rv)",
     xlab = "Iteration (t)", ylab = "Value")

# Add a red dashed line to mark the end of burn-in
abline(v = 2000, col = "red", lty = 2, lwd = 2)
legend("topright", legend = "End of Burn-in", col = "red", lty = 2)
```

# 4.2 Multiple chains (running the algorithm multiple times in parallel)

```{r}
library(parallel)


# function to run chains in parallel (Mac/Linux use mclapply; Windows requires parLapply)
chains_list <- mclapply(1:n_chains, function(i) {
  gibbs_har(y, X, n_iter = 10000, burnin = 2000)
}, mc.cores = n_chains)


# Run 2 chains (i.e. running the MCMC algorithm twice)
n_chains <- 2

# lapply(): applies a function over a list or vector
chains_out <- lapply(1:n_chains, function(i) {
  # Start from a different random beta 
  # This helps verify if they converge to the same posterior
  beta_init <- rnorm(ncol(X), 0, 1) 
  
  # Call your function (make sure it returns the full n_iter)
  gibbs_har(y, X, n_iter = 10000, burnin = 2000)
})


str( chains_out ) # displaying the structure of the list object

# comparing the posterior colMeans of both chains
colMeans(chains_out[[1]]$beta)
colMeans(chains_out[[2]]$beta)


```


